Seagrass mapping and monitoring \\ along the coasts of Crete, Greece

### 2011-02-11 MScDiplom_Lemenkova.tex
Valentijn's new comments (11 Feb). From total 36 finished 14, 15, 20, 23, 29, 34. Partly done: 18, 19.

\subsection{Research objective}\index[concepts]{Research!objective}
The current MSc research aims to explore the environmental conditions for the spatial distribution of \textit{P.oceanica} seagrass along the northern coast of Crete Island, based on the remote sensing and \ac{GIS} techniques, knowledge about the coastal environment in Crete and integration of various data
from the following sources \index[concepts]{Data}:
	\begin{itemize}
		\item[] i) spectra of \textit{P.oceanica}, carbonate sand, silt and other seafloor types
		\item[] ii) satellite imagery: \ac{Landsat TM}, \ac{Landsat ETM+}
		\item[] iii) aerial photos: Google Earth
		\item[] iv) \textit{in-situ} fieldwork data of underwater videographic measurements
		\item[] v) vector GIS layers.
	\end{itemize}
The research aims to explore the limitations of the application of the \ac{RTM} and remote sensing techniques 
towards the study of the environmental properties of the \textit{P.oceanica} meadows distribution
over the selected locations of northern Crete. The main research objective is monitoring the seagrass \textit{P.oceanica} in selected areas of Crete and analyzing the environmental conditions for the growth of \textit{P.oceanica}. \\
This research is supported by the \textit{in-situ} measurements in two selected locations of the northern coast
of Crete (Ligaria beach in Agia Pelagia district and Xerocampos), using following methods of the
remote sensing techniques: spectral modelling by means of the \ac{RTM} \ac{WASI}, underwater videometric measurements made by \ac{Olympus ST} camera, aerial Google Earth and satellite images from different sources, spatial \ac{GIS} and statistical analysis.

@@

properties of seawater.

@@

@@

\begin{figure}[H]
	\centering
	\includegraphics[scale=0.5]{GNU-17.jpg}
	\caption{Optical properties of the sea water with sediments, measured in aquarium tank. Agia Pelagia district, Crete. \href{http://www.gnuplot.info/}				{Gnuplot}. Pages: $<$\pageref{page-21}$>$, $<$\pageref{page-31}$>$}\label{fig:4.2}
\end{figure}

@@

In order to focus on the factors of primary importance, other and less influencing factors are excluded,
i.e. sun angle, concentration of the suspected particles in the water column, content of gelbstoff, etc.
For these factors default values of \ac{WASI} are accepted.
Under normal conditions by independent water colour sampling, values of the remote-sensing
reflectance can vary by 12 - 24 \% \cite{Toole00}\label{Toole00}, and these variations in the radiometric
determinations are mainly caused by the variety of the environmental factors.
Different factors influence colour and spectral reflectance of water, among which different bottom
depths, concentration of suspended particles in water column, water temperature, sun angle,
concentration of \ac{CDOM}, concentration of phytoplankton,
aerosol scattering, exponent of backscattering by small particles, cloudiness, viewing geometry and
wind speed (which is, however, not the major source of uncertainty). All these environmental
components increase the absorption and scattering of light which, in its turn, results in a complex
relationship between their concentrations and the radiance of water that finally influence its spectral
reflectance.\\
However, although there are a variety of environmental factors that contribute to the spectral
reflectance, the most important ones from the mentioned above are the water column (depth) and seafloor fraction.
It is because spectra of the seagrass \textit{P.oceanica} vary qualitatively over the depths interval of 0.5–4 m,
and secondly, the content and cover fraction of the seafloor have the most distinctive effect on the
spectral reflectance of the water. \\
Therefore, in the current research we focus on these two major factors, and study the response of the
water reflectance towards changing conditions of the water column depth and seafloor bottom cover
fraction (seagrass \textit{P.oceanica}, carbonate sand and silt).\\
The analysis of the spectral signatures of the seagrass \textit{P.oceanica} and sand clearly shows (Fig.\ref{fig:27}) that seagrass
has spectral reflectance much lesser than that of a carbonate sand, in general not increasing values of
10 \% reflectance in spectra of 500-600 nm, while sand has spectral reflectance approaching 33\% in its
highest values.

\subsubsection{Water color spectral simulation using WASI}\index[concepts]{Data!processing}\label{page-35}
The main aim of this part of research is to clarify if the bottom reflectance of the different seafloor types including patches of
the seagrass \textit{P.oceanica} meadows, silt and carbonate sand differ and can be clearly discriminated while mapping. A study is
based on three seafloor types containing silt, carbonate sand and seagrass, as well as mixed
types, where the spectral signatures were examined. \ac{WASI} software\index[concepts]{Seawater optical properties!RTM!WASI} (Fig.\ref{fig:4.7}) is used to simulate
spectral reflectance and colour discrimination of water, affected by presence of \textit{P.oceanica} and other factors, under various
environmental conditions which influence its colour.

\begin{figure}[h]
	\centering
	\includegraphics[scale=0.30]{Fig-30.jpg}
	\caption{Main interface of \ac{WASI} water color spectral simulation: \\ Spectral reflectance of \textit{P.oceanica} and silt at 0.5-4.0m. Pages: $<$\pageref{page-35}$>$, $<$\pageref{page-31}$>$}
	\label{fig:4.7}
\end{figure}

The data provided with the model was determined at freshwater of the Boddensee \cite{Gege05}\label{Gege05}, 
yet the model was adjusted for the marine environment, so that its parameters (concentration of chlorophyll, concentration of small particles,
yellow substance, etc) now perfectly simulate the Mediterranean water conditions.
The remote sensing reflectance has been compared under the conditions of different water depths and
cover fraction of the seafloor, in order to assess spectral signatures of the seagrass and carbonate sand
as major seafloor types. \ac{WASI} enables to use forward or inverse calculation for the spectrum types at
a diapason of 350-800 nm with a 1nm spectral resolution. For the spectral analysis we applied
forward calculations, i.e. a computing and plotting of series of spectra according to specified
parameter settings, with exactly defined depths and cover fraction. \\
The specific parameters have been chosen for the simulation of the environmental conditions where seagrass
grow. The adaptations to life in salt sea water requires various physical and chemical parameters
which include salinity, temperature of $17-25\,^{\circ}C$, light requirements with 10-20 \% on average, ranging
from 4.4 \%  minimal up to 29 \% \cite{McKenzie09}\label{McKenzie09}, so that the zenith angle is taken as $35-45\,^{\circ}$
and reflection factor 0.0201. \textit{P.oceanica}\label{page-36}. These values, simulating the environment of the Mediterranean Sea, are fixed (Tab.\ref{tab:3}) among WASI user-defined parameters.
The calculations are done for the spectrum 350-800 nm, covering the most important part of the RS spectrum: 1) Blue-green 0.45 - 0.5 μm; 2) Green 0.5 - 0.6 μm; 3) Red 0.6 - 0.7 μm; 4) Red-NIR 0.7 - 0.8 μm (Fig.\ref{fig:4.8}).

\begin{figure}[h]
	\centering
	\subfloat [Concentration of Gelbstoff at depths 0.5 – 8.0 m]{\label{fig:31a}\includegraphics[width=0.4\textwidth]{Fig-31a.jpg}}
	\subfloat [Bottom reflectance at depth 0.5-4.0 meters]{\label{fig:31b}\includegraphics[width=0.4\textwidth]{Fig-31b.jpg}}
	\caption{Sea water physical properties, modeled by \ac{WASI}. Page: $<$\pageref{page-36}$>$}
	\label{fig:4.8}
\end{figure}

\paragraph{Model parameters: depth and bottom cover fraction}
Although seagrass \textit{P.oceanica} can be found until depth limits down to 40 m depth (Hartog
1970), the most preferable limits of its distribution in the Mediterranean Sea are shallow waters until
4 meters of depth.

\begin{table}
	\rowcolors{1}{Aquamarine}{Azure}
	\caption{Model-specific parameters of water WASI adjusted to simulate environment of the Mediterranean Sea along Crete. Page: $<$\pageref{page-36}$>$}
	\centering
	 \begin{tabular}{|  p{2cm} | p{6cm} | p{25mm} | }
	    \hline
		    \textbf{Parameter, WASI} & \textbf{Name and description} & \textbf{Values} \\ \hline\hline
		    C-L & Concentration of large suspended particles & 8 \\ \hline
		   C(i), i=0..5 & Concentration of Phytoplankton & 0.035 -0.089 $ug^{-1}$ \\ \hline
		    bbS & Specific backscattering for small particles & 0.005 $m^2g^{-1}$ \\ \hline
		   T-W & Temperature of water & $17-25\,^{\circ}C$ \\ \hline
		    n & Exponent of Backscattering by small particles & 0.005 $m^2g^{-1}$ \\ \hline
		    Q &Anisotropy factor of upwelling radiation ("Q-factor") & 5.00 \\ \hline
		    sigma-L & Reflection factor of sky radiance & 0.0201 \\ \hline
		    b1 & Backscattering coefficient of saline waters & 0.00144 $m^{-1}$ \\ \hline
		    λ0 & Reference Wavelength for \ac{CDOM} (Gelbstoff) absorption & 440 \\ \hline
		    sun & Sun zenith angle & 45.0 \\ \hline
		    zB & Bottom depth & 4.00 \\ \hline
		   f(i), i=0...5 & Areal fraction of bottom surface type number n & 01/10/10 \\ \hline
		    K0 & Coefficient of Attenuation & 1.0546 \\ \hline
		   view & Viewing angle (0 = nadir) & 0\\ \hline
		    C-X & Concentration of non-chlorophyll particles (absorption at $\lambda_0$) & 0 \\ \hline
		   λS & Reference wavelength for scattering of small particles & 500 \\ \hline
		    C-S & Concentration of small suspended particles & 0 \\ \hline
		   S & Exponent of \ac{CDOM} absorption & 0.0140 \\ \hline
		    C-Y & Concentration of \ac{CDOM} Gelbstoff (absorption at $\lambda_0$) & 0.400 \\ \hline
		    Bn & BDRF of bottom reflectance (sand) & 0.318 $sr^{-1}$\\
	    \hline
	  \end{tabular}
	   \label{tab:3}
\end{table}

@@

An analysis of the spectral reflectance of \textit{P.oceanica} is done using the \ac{WASI} simulations\index[concepts]{Seawater optical properties!RTM!WASI} in order to determine, which wavebands can be still used to identify \textit{P.oceanica}. The analysis of spectra shows
that the appropriate wavebands for seagrass mapping lay between 500 and 600 nm and has
also peaks at around 700 nm, ca between 680 and 710 nm (Fig.\ref{fig:4.10}). The highest values of the bottom
reflectance are at spectra of 500-600 nm. The most appropriate depths at which the spectral signatures of the seagrass
could still be discriminated are lesser than 2.5 meters.
The remote sensing properties of the broad-band satellite sensors are often limited and do not allow
easy and unique discrimination of seagrass from other seafloor types. Therefore, we focus on narrow-band
spectra with the resolution of 1 nm.

\paragraph{Spectral discrimination of \textit{P.oceanica}}\label{page-39}
The patches of white sandy bottoms of the seafloor, studied in the fieldwork in Ligaria beach, are
much brighter than mattes of \textit{P.oceanica} (Fig.\ref{fig:A.6}, \ref{fig:A.5}): seafloor types in Ligaria), which can be clearly seen
at the graph comparing values of the spectral reflectance of the carbonate sand (Fig.\ref{fig:A.35}) and that of \textit{P.oceanica} (Fig.\ref{fig:A.28}).
The graph is received in excel spreadsheet using mean values of spectral reflectance of
sand (Fig.\ref{fig:A.27}) and seagrass (Fig.\ref{fig:A.35}), respectively, which have been calculated from measurements of radiance and
irradiance received in Agia Pelagia bay.

\begin{figure}[H]
	\centering
	\includegraphics[scale=0.40]{GNU-19.jpg}
	\caption{Statistical analysis of the spectral reflectance of \textit{P.oceanica}: min-max, average values (red bold points), Q1-Q3 areas (green vertical dashed) and 	measured values (dotted): multiplot of measurement sets 200-300. \href{http://www.gnuplot.info/}{Gnuplot}. Pages: $<$\pageref{page-21}$>$, $<$\pageref{page-28}$>$, $<$\pageref{page-37}$>$}
	\label{fig:4.10}
\end{figure}
\begin{figure}[H]
	\centering
	\subfloat [Sand]{\label{fig:34a}\includegraphics[width=0.4\textwidth]{Fig-27a.jpg}}
	\subfloat [\textit{P.oceanica}]{\label{fig:34b}\includegraphics[width=0.4\textwidth]{Fig-27b.jpg}}
	\caption{Bottom albedo of carbonate sand and \textit{P.oceanica}, Agia Pelagia. Page: $<$\pageref{page-22}$>$}
	\label{fig:4.11}
\end{figure}

@@

\paragraph{ANOVA-testing}\index[concepts]{Data!statistical analysis!ANOVA}
We applied \ac{ANOVA} (ANalysis Of Variance) testing for the analysis of the probability of the
spectral reflectance of \textit{P.oceanica} to be more or less spectrally distinct from other seafloor cover
types with changing depth. W are checking the statement Hypothesis $H_o$, is true, which is “the
spectral distinguishability of the seagrass \textit{P.oceanica} from other seafloor types is not changing with
varying \textit{in-situ} conditions, $H_o$: μ1 =μ2 =μ3=...= μn.” which gives the following outcome.
P(reject $H_o$|$H_o$ is valid)= P(X$>$c|p=)=.05
where c comes for critical values and p=probability.
The result P=.05 is (Tab.\ref{tab:8}), hence, very small, which makes the statement of Hypothesis $H_o$ highly unlikely
(less than 1 in a 10 chance). The one-way \ac{ANOVA} highlighted a significant difference between data
of spectral reflectances of \textit{P.oceanica} at different depths (Tab.\ref{tab:9}), i.e. true is the opposite statement:
Hypothesis $H_a$=“the spectral discernibility of the seagrass \textit{P.oceanica} is distinctly changing with
varying \textit{in-situ} conditions, i.e. increasing depth, $H_o$: μ$\neq$μ2$\neq$μ3$\neq$ ... $\neq$μn“.

\paragraph{Student-t testing}\index[concepts]{Data!statistical analysis!Student-t testing}
The Student-t test, one of the most commonly used techniques for testing a hypothesis on the basis
of a difference between sample means, is used for the data analysis. In our case the Student t-test demonstrates, if the variation
between two analyzed groups – values of spectral reflectance of seagrass \textit{P.oceanica} and sand - is
significant. Therefore, we use Student t-test to compare two sets of quantitative data of spectral
reflectance of \textit{P.oceanica} and sand, respectively, with samples collected independently of one
another. The Student-t test can be performed knowing just the means, standard deviation, and number
of data points. Therefore, we used the data (\nameref{app}) of means of spectral reflectance of the both
 cover types within data sets, their standard deviation, and the number of data points.
The results show the following outcome.
As a result of the statistical testing, we came to the following conclusion.
The Hypothesis $H_a$ is true, which claims that the spectral discernibility of the seagrass \textit{P.oceanica} is
distinctly changing with varying \textit{in-situ} conditions, i.e. increasing depth, $H_o$: μ1≠μ2≠μ3≠...≠ μn .
Statistical results are illustrated with Error bars made using Excel's vertical box and Whisker Charts
(Box Plots).
The graphs were plotted on the basis of the following statistical data calculated from the sampling
measurements data: Table 2.

@@

\subsection{Image processing using Erdas Imagine}\label{page-47}
The analysis of the imagery of Cretan coasts is based on the images classification and is aimed to
investigate the distribution of the seagrass \textit{P.oceanica} within the research area.
In the image processing part of the research supervised classification has been applied to the aerial
and satellite images. Seagrass meadows and other seafloor cover types were evaluated through a
detailed examination of the imagery. \\
Seagrass beds are clearly visible in color aerial Google Earth\index[concepts]{Remote sensing!aerial imagery!GoogleEarth}
photographs (Fig.\ref{fig:A.36}, Fig.\ref{fig:A.37}), contrasting against slightly-yellow and brownish sand bottom. The seagrass areas are
detected using different bands combinations, masked and studied for the estimation of the changes in
the areas. The classification is based on the spectral properties of the \textit{P.oceanica}, such as brightness,
colour, texture and structure of the seagrass mattes (Fig.\ref{fig:38}a). The raster-based mapping\label{page-48}
includes supervised classification with training sites of seagrasses (10-15 set areas) in different bands
for each photograph by classification a series of polygons characteristic of each of the sea floor
bottom types: sandy surface, seagrass bed for each species including \textit{P.oceanica} (meadows), patchy
seagrass bed and algae on rock, rocks, muddy surface, etc (Fig.\ref{fig:38}b).

\subsection{Seagrass mapping using ArcGIS}

\subsubsection{Data integration}\index[concepts]{Data!integration}
The integrated approach used in this research work has high potential as a means to monitor changes
in seagrass landscape occurring in shallow waters over Crete area.

@@

\subsubsection{Upscale GIS mapping of the seagrass landscapes}\index[concepts]{Seagrasses!spatial structure}\label{page-48}
Although ocean pelagic landscapes have a high degree of spatial variance and less structurally
complex, comparing to the terrestrial ones, the landscape-level phenomena have similar features, and
there are accepted definitions of landscapes elements within the seagrass meadows \cite{Robbins94}\label{Robbins94}. 
In general, the structure of seagrass landscape is simpler than that of the terrestrial
ecosystems in biodiversity and complexity; however, seagrass landscapes show variation in spatial
patterns over different special scales. The complexity of the landscape of seagrass meadow is shown by the
measure of patches in size and shape, expressed in ratio of patch perimeter to area. The fragmentation
of the seagrass landscapes is expressed in contiguity displaying patch aggregation within meadows.\\
The general principles of the hierarchy within the seagrass landscapes are based upon the quantitative
analysis of the spatial patterns, consisted by components and separate elements. Thus, bunches of
individual shoots construct patches, the first hierarchical level. Patches are arranged into discrete
clumps of mattes (at a scale of centimeters to meters) which, in turn, make up beds with l-100m in
diameter. Finally, seagrass beds are arranged into meadows that may extend over kilometer-wide
areas, historically defined as landscapes \cite{Robbins94}\label{Robbins94}.

Besides spatial structure of the seagrass meadows, there are strong and complex patterns of depth zoning, specific to individual
seagrass species, but such detailed classification goes beyond the scope of the current MSc research.
Unique feature of seagrass meadows, most characteristic for shallow areas, is dynamics of their
landscapes. Homogeneous, continuous seagrass meadows can reach in size up to 40 m2 (Fig.\ref{fig:38} a),
however they are often interrupted by gaps and channels of open spaces generated by the complex,
disturbing sedimentation processes, turbulence and turbidity regime of waters and landscape dynamics,
which form gaps and channels within meadows, making them “patchy-looking”, i.e. diversified by
separate mattes (Fig.\ref{fig:38} b). The traditional definition for gap in vegetation cover is “disturbance generated
openings in either floral or faunal cover” \cite{Connell78}\label{Connell78}.
Formation and increase of these
gaps within the mosaic of seagrass meadows is caused by different reasons. The most probable
drivers for the process of gaps formation within seagrass meadows are removal of interior vegetation,
differential growth of seagrass meadows and increased sedimentation. Thus, storms lead to severe
deposition of sediments, burying parts of seagrass meadow, the same effect has movements and
deposition of sediments during and after floods\cite{Bell99}\label{Bell99}.
Increased nutrient sedimentation, especially phosphorus, were explored \cite{Jensen01}\label{Jensen01} as a
potential mechanism for increasing patch dynamics and morphological plasticity within seagrass
meadows. Finally, increasing the degree of fragmentation of the landscapes of 	extit{P.oceanica} meadows
can be caused by the invasion of alien species, such as \textit{Cymodocea nodosa, Caulerpa prolifera,
Caulerpa taxifolia}. Invaders are in general strong colonizers comparing to native \textit{P.oceanica}: they
occupy much greater habitat space within the regressed meadows of stressed native seagrass
\cite{Montefalcone10}\label{Montefalcone10}. However, on the northern coasts of Crete the only dominating seagrass
species is \textit{P.oceanica}.
Morphological differences in scale of seagrass landscape formations, discussed above, cause need for
the different-scale mapping. Therefore, the investigation of the seagrass meadows at different levels is
performed using underwater videometric measurements, aerial and satellite imagery.

%%%%%%%%%%%%%%%%%%%%%%%%% Discussion %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}\label{sec:5}

>>>>

\subsection{Research objective}\index[concepts]{Research!objective}
The current MSc research aims to explore the environmental conditions for the spatial distribution of \textit{P.oceanica} seagrass along the northern coast of Crete Island, based on the remote sensing and \ac{GIS} techniques, knowledge about the coastal environment in Crete and integration of various data
from the following sources \index[concepts]{Data}:
	\begin{itemize}
		\item[] i) spectra of \textit{P.oceanica}, carbonate sand, silt and other seafloor types
		\item[] ii) satellite imagery: \ac{Landsat TM}, \ac{Landsat ETM+}
		\item[] iii) aerial photos: Google Earth
		\item[] iv) \textit{in-situ} fieldwork data of underwater videographic measurements
		\item[] v) vector GIS layers.
	\end{itemize}
Although there are a variety of environmental factors that contribute to the spectral
reflectance, the most important ones are water column height and seafloor fraction.
It is because spectra of the seagrass \textit{P.oceanica} vary qualitatively over the depths interval of 0.5–4 m,
and secondly, the content and cover fraction of the seafloor have the most distinctive effect on the
spectral reflectance of the water. \\
Therefore, in the current research we focus on these two major factors, and study the response of the
water reflectance towards changing conditions of the water column depth and seafloor bottom cover
fraction (seagrass \textit{P.oceanica}, carbonate sand and silt).\\
The research aims to explore the limitations of the application of the \ac{RTM} and remote sensing techniques 
towards the study of the environmental properties of the \textit{P.oceanica} meadows distribution
over the selected locations of northern Crete. The main research objective is monitoring the seagrass \textit{P.oceanica} in selected areas of Crete and analyzing the environmental conditions for the growth of \textit{P.oceanica}. \\
This research is supported by the \textit{in-situ} measurements in two selected locations of the northern coast
of Crete (Ligaria beach in Agia Pelagia district and Xerocampos), using following methods of the
remote sensing techniques: spectral modelling by means of the \ac{RTM} \ac{WASI}, underwater videometric measurements made by \ac{Olympus ST} camera, aerial Google Earth and satellite images from different sources, spatial \ac{GIS} and statistical analysis.

@@

properties of seawater.\\
We used several statistical methods to asses data received as a result of measurements. 
The Student-t test\index[concepts]{Data!statistical analysis!Student-t testing},
one of the most commonly used techniques for testing a hypothesis on the basis
of a difference between sample means, is used for the data analysis. In our case the Student t-test demonstrates, if the variation
between two analyzed groups – values of spectral reflectance of seagrass \textit{P.oceanica} and sand - is
significant. Therefore, we use Student t-test to compare two sets of quantitative data of spectral
reflectance of \textit{P.oceanica} and sand, respectively, with samples collected independently of one
another. The Student-t test can be performed knowing just the means, standard deviation, and number
of data points. Therefore, we used the data (\nameref{app}) of means of spectral reflectance of the both
 cover types within data sets, their standard deviation, and the number of data points.

@@

\subsection{Image processing using Erdas Imagine}\label{page-47}
The analysis of the imagery of Cretan coasts is based on the images classification and is aimed to
investigate the distribution of the seagrass \textit{P.oceanica} within the research area.
In the image processing part of the research supervised classification has been applied to the aerial
and satellite images. Seagrass meadows and other seafloor cover types were evaluated through a
detailed examination of the imagery. \\
Seagrass beds are clearly visible in color aerial Google Earth\index[concepts]{Remote sensing!aerial imagery!GoogleEarth}
photographs (Fig.\ref{fig:A.36}, Fig.\ref{fig:A.37}), contrasting against slightly-yellow and brownish sand bottom. The seagrass areas are
detected using different bands combinations, masked and studied for the estimation of the changes in
the areas. The classification is based on the properties of the \textit{P.oceanica}, such as brightness,
colour, texture and structure of the seagrass mattes (Fig.\ref{fig:38}a). The raster-based mapping\label{page-48}
includes supervised classification with training sites of seagrasses (10-15 set areas) in different bands
for each photograph by classification a series of polygons characteristic of each of the sea floor
bottom types: sandy surface, seagrass bed for each species including \textit{P.oceanica} (meadows), patchy
seagrass bed and algae on rock, rocks, muddy surface, etc (Fig.\ref{fig:38}b).

@@

\begin{figure}[H]
	\centering
	\includegraphics[scale=0.5]{GNU-17.jpg}
	\caption{Optical properties of the sea water with sediments, measured in aquarium tank. Agia Pelagia district, Crete. \href{http://www.gnuplot.info/}{Gnuplot}. Pages: $<$\pageref{page-21}$>$, $<$\pageref{page-31}$>$}\label{fig:4.2}
\end{figure}

@@

In order to focus on the factors of primary importance, other and less influencing factors are excluded,
i.e. sun angle, concentration of the suspected particles in the water column, content of gelbstoff, etc.
For these factors default values of \ac{WASI} are accepted.
Under normal conditions by independent water colour sampling, values of the remote-sensing
reflectance can vary by 12 - 24 \% \cite{Toole00}\label{Toole00}, and these variations in the radiometric
determinations are mainly caused by the variety of the environmental factors.
Different factors influence colour and spectral reflectance of water, among which different bottom
depths, concentration of suspended particles in water column, water temperature, sun angle,
concentration of \ac{CDOM}, concentration of phytoplankton,
aerosol scattering, exponent of backscattering by small particles, cloudiness, viewing geometry and
wind speed (which is, however, not the major source of uncertainty). All these environmental
components increase the absorption and scattering of light which, in its turn, results in a complex
relationship between their concentrations and the radiance of water that finally influence its spectral
reflectance.

\subsubsection{Water color spectral simulation using WASI}\index[concepts]{Data!processing}\label{page-35}
The main aim of this part of research is to clarify if the bottom reflectance of the different seafloor types including patches of
the seagrass \textit{P.oceanica} meadows, silt and carbonate sand differ and can be clearly discriminated while mapping. A study is
based on three seafloor types containing silt, carbonate sand and seagrass, as well as mixed
types, where the spectral signatures were examined. \ac{WASI} software\index[concepts]{Seawater optical properties!RTM!WASI} (Fig.\ref{fig:4.7}) is used to simulate
spectral reflectance and colour discrimination of water, affected by presence of \textit{P.oceanica} and other factors, under various
environmental conditions which influence its colour.
The data provided with the model was determined at freshwater of the Boddensee \cite{Gege05}\label{Gege05}, 
yet the model was adjusted for the marine environment, so that its parameters (concentration of chlorophyll, concentration of small particles,
yellow substance, etc) now perfectly simulate the Mediterranean water conditions.
The remote sensing reflectance has been compared under the conditions of different water depths and
cover fraction of the seafloor, in order to assess spectral signatures of the seagrass and carbonate sand
as major seafloor types. \ac{WASI} enables to use forward or inverse calculation for the spectrum types at
a diapason of 350-800 nm with a 1nm spectral resolution. For the spectral analysis we applied
forward calculations, i.e. a computing and plotting of series of spectra according to specified
parameter settings, with exactly defined depths and cover fraction.

\begin{figure}[H]
	\centering
	\subfloat [Concentration of Gelbstoff at depths 0.5 – 8.0 m]{\label{fig:31a}\includegraphics[width=0.4\textwidth]{Fig-31a.jpg}}
	\subfloat [Bottom reflectance at depth 0.5-4.0 meters]{\label{fig:31b}\includegraphics[width=0.4\textwidth]{Fig-31b.jpg}}
	\caption{Sea water physical properties, modeled by \ac{WASI}. Page: $<$\pageref{page-36}$>$}
	\label{fig:4.8}
\end{figure}

The specific parameters have been chosen for the simulation of the environmental conditions where seagrass
grow. The adaptations to life in salt sea water requires various physical and chemical parameters
which include salinity, temperature of $17-25\,^{\circ}C$, light requirements with 10-20 \% on average, ranging
from 4.4 \%  minimal up to 29 \% \cite{McKenzie09}\label{McKenzie09}, so that the zenith angle is taken as $35-45\,^{\circ}$
and reflection factor 0.0201. \textit{P.oceanica}\label{page-36}. These values, simulating the environment of the Mediterranean Sea, are fixed (Tab.\ref{tab:3}) among WASI user-defined parameters.
The calculations are done for the spectrum 350-800 nm, covering the most important part of the RS spectrum: 1) Blue-green 0.45 - 0.5 μm; 2) Green 0.5 - 0.6 μm; 3) Red 0.6 - 0.7 μm; 4) Red-NIR 0.7 - 0.8 μm (Fig.\ref{fig:4.8}).

\paragraph{Model parameters: depth and bottom cover fraction}
Although seagrass \textit{P.oceanica} can be found until depth limits down to 40 m depth \cite{DenHartog70}\label{DenHartog70}, the most preferable limits of its distribution in the Mediterranean Sea are shallow waters until
4 meters of depth.

\begin{table}[H]\footnotesize
	\rowcolors{1}{Aquamarine}{Azure}
	\caption{Model-specific parameters of water WASI adjusted to simulate environment of the Mediterranean Sea along Crete. Page: $<$\pageref{page-36}$>$}
	\centering
	 \begin{tabular}{|  p{2cm} | p{6cm} | p{25mm} | }
	    \hline
		    \textbf{Parameter, WASI} & \textbf{Name and description} & \textbf{Values} \\ \hline\hline
		    C-L & Concentration of large suspended particles & 8 \\ \hline
		   C(i), i=0..5 & Concentration of Phytoplankton & 0.035 -0.089 $ug^{-1}$ \\ \hline
		    bbS & Specific backscattering for small particles & 0.005 $m^2g^{-1}$ \\ \hline
		   T-W & Temperature of water & $17-25\,^{\circ}C$ \\ \hline
		    n & Exponent of Backscattering by small particles & 0.005 $m^2g^{-1}$ \\ \hline
		    Q &Anisotropy factor of upwelling radiation ("Q-factor") & 5.00 \\ \hline
		    sigma-L & Reflection factor of sky radiance & 0.0201 \\ \hline
		    b1 & Backscattering coefficient of saline waters & 0.00144 $m^{-1}$ \\ \hline
		    λ0 & Reference Wavelength for \ac{CDOM} (Gelbstoff) absorption & 440 \\ \hline
		    sun & Sun zenith angle & 45.0 \\ \hline
		    zB & Bottom depth & 4.00 \\ \hline
		   f(i), i=0...5 & Areal fraction of bottom surface type number n & 01/10/10 \\ \hline
		    K0 & Coefficient of Attenuation & 1.0546 \\ \hline
		   view & Viewing angle (0 = nadir) & 0\\ \hline
		    C-X & Concentration of non-chlorophyll particles (absorption at $\lambda_0$) & 0 \\ \hline
		   λS & Reference wavelength for scattering of small particles & 500 \\ \hline
		    C-S & Concentration of small suspended particles & 0 \\ \hline
		   S & Exponent of \ac{CDOM} absorption & 0.0140 \\ \hline
		    C-Y & Concentration of \ac{CDOM} Gelbstoff (absorption at $\lambda_0$) & 0.400 \\ \hline
		    Bn & BDRF of bottom reflectance (sand) & 0.318 $sr^{-1}$\\
	    \hline
	  \end{tabular}
	   \label{tab:3}
\end{table}

@@

An analysis of the spectral reflectance of \textit{P.oceanica} is done using the \ac{WASI} simulations\index[concepts]{Seawater optical properties!RTM!WASI} in order to determine, which wavebands can be still used to identify \textit{P.oceanica}. The analysis of spectra shows
that the appropriate wavebands for seagrass mapping lay between 500 and 600 nm and has
also peaks at around 700 nm, ca between 680 and 710 nm (Fig.\ref{fig:4.10}). The highest values of the bottom
reflectance are at spectra of 500-600 nm. The most appropriate depths at which the spectral signatures of the seagrass
could still be discriminated are lesser than 2.5 meters.

\paragraph{Spectral discrimination of \textit{P.oceanica}}\label{page-39}
The patches of white sandy bottoms of the seafloor, studied in the fieldwork in Ligaria beach, are
much brighter than mattes of \textit{P.oceanica} (Fig.\ref{fig:A.6}, \ref{fig:A.5}): seafloor types in Ligaria), which can be clearly seen
at the graph comparing values of the spectral reflectance of the carbonate sand (Fig.\ref{fig:A.35}) and that of \textit{P.oceanica} (Fig.\ref{fig:A.28}).

\begin{figure}[H]
	\centering
	\includegraphics[scale=0.40]{GNU-19.jpg}
	\caption{Statistical analysis of the spectral reflectance of \textit{P.oceanica}: min-max, average values (red bold points), Q1-Q3 areas (green vertical dashed) and 	measured values (dotted): multiplot of measurement sets 200-300. \href{http://www.gnuplot.info/}{Gnuplot}. Pages: $<$\pageref{page-21}$>$, $<$\pageref{page-28}$>$, $<$\pageref{page-37}$>$}
	\label{fig:4.10}
\end{figure}
\begin{figure}[H]
	\centering
	\subfloat [Sand]{\label{fig:34a}\includegraphics[width=0.3\textwidth]{Fig-27a.jpg}}
	\subfloat [\textit{P.oceanica}]{\label{fig:34b}\includegraphics[width=0.3\textwidth]{Fig-27b.jpg}}
	\caption{Bottom albedo of carbonate sand and \textit{P.oceanica}, Agia Pelagia. Page: $<$\pageref{page-22}$>$}
	\label{fig:4.11}
\end{figure}
The graph is received in excel spreadsheet using mean values of spectral reflectance of
sand (Fig.\ref{fig:A.27}) and seagrass (Fig.\ref{fig:A.35}), respectively, which have been calculated from measurements of radiance and
irradiance received in Agia Pelagia bay.

@@

\paragraph{ANOVA-testing}\index[concepts]{Data!statistical analysis!ANOVA}
We applied \ac{ANOVA} (ANalysis Of Variance) testing for the analysis of the probability of the
spectral reflectance of \textit{P.oceanica} to be more or less spectrally distinct from other seafloor cover
types with changing depth. W are checking the statement Hypothesis $H_o$, is true, which is “the
spectral distinguishability of the seagrass \textit{P.oceanica} from other seafloor types is not changing with
varying \textit{in-situ} conditions, $H_o$: μ1 =μ2 =μ3=...= μn.” which gives the following outcome.
P(reject $H_o$|$H_o$ is valid)= P(X$>$c|p=)=.05
where c comes for critical values and p=probability.
The result P=.05 is (Tab.\ref{tab:8}), hence, very small, which makes the statement of Hypothesis $H_o$ highly unlikely
(less than 1 in a 10 chance). The one-way \ac{ANOVA} highlighted a significant difference between data
of spectral reflectances of \textit{P.oceanica} at different depths (Tab.\ref{tab:9}), i.e. true is the opposite statement:
Hypothesis $H_a$=“the spectral discernibility of the seagrass \textit{P.oceanica} is distinctly changing with
varying \textit{in-situ} conditions, i.e. increasing depth, $H_o$: μ$\neq$μ2$\neq$μ3$\neq$ ... $\neq$μn“.
As a result of the statistical testing, we came to the following conclusion.
The Hypothesis $H_a$ is true, which claims that the spectral discernibility of the seagrass \textit{P.oceanica} is
distinctly changing with varying \textit{in-situ} conditions, i.e. increasing depth, $H_o$: μ1≠μ2≠μ3≠...≠ μn .
Statistical results are illustrated with Error bars made using Excel's vertical box and Whisker Charts
(Box Plots).
The graphs were plotted on the basis of the following statistical data calculated from the sampling
measurements data: Table 2.

@@

\subsection{Integrated mapping of seagrass}

\index[concepts]{Data!integration}
The integrated approach used in this research work has high potential as a means to monitor changes
in seagrass landscape occurring in shallow waters over Crete area.

@@

%%%%%%%%%%%%%%%%%%%%%%%%% Discussion %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}\label{sec:5}

\subsection{Remote sensing for seagrass mapping}
